---
title: "Statistical Inference Course Project"
author: "Chris van Hasselt"
date: "Monday, May 11, 2015"
output:
  pdf_document:
    fig_height: 3.75
    fig_width: 5
    latex_engine: xelatex
---
```{r libraries, echo=FALSE }
# libraries loaded here
library(ggplot2)
```

# Overview:
The exponential distribution can be simulated in R with *rexp(n, lambda)* where *lambda* is the rate parameter. The mean of exponential distribution is *1/lambda* and the standard deviation is also also *1/lambda*. In this report, I will investigate the distribution of averages of 40 exponentials, when the rate parameter *lambda = 0.2*.

Specifically, I illustrate the properties of the distribution of the mean of 40 exponentials, showing:

#. The sample mean and compare it to the theoretical mean of the distribution;
#. variability (via variance) with respect to the theoretical variance of the distribution;
#. that the distribution is approximately normal.

Plots generated in this document are created using the *ggplot2* library.

# Simulations:
To analyze empirical versus theoretical statistics, I create a data.frame named *dfSamples* of random exponential values based on the given rate, *lambda = 0.2*.  Code available in the Appendix.

```{r introduction,echo=FALSE}

set.seed(2132)          # set random seed, for repeatability

lambda <- 0.2           # define lambda, the exponential rate

numSimulations <- 1000  # numSimulations = number of simulations (1000)

sampleSize <- 40        # sampleSize = number of exponentials per simulation (40)

# allSamples is the numSimulations (1000) times sampleSize (40) exponentials
allSamples <- rexp(sampleSize*numSimulations, rate=lambda)

# sampleMatrix packages the exponentials as a matrix 
sampleMatrix <- matrix(allSamples, numSimulations)

# dfSamples is a data.frame of all the samples
dfSamples   <- as.data.frame(sampleMatrix)

```
# Sample Mean versus Theoretical Mean:
To compare the empirical mean to the the theoretical mean, I calculate row means for data.frame *dfSamples* as *xbar*, and cumulative row means as *xbarCumulative*.  The plot below shows the means of samples of 40 exponentials converging toward the theoretical mean of *1/lambda*, in this case 5, demonstrating asymptotic convergence to the theoretical mean.
```{r compare_mean, echo=FALSE}

# dfSamples$xbar is a column appended of row means
dfSamples$xbar <- apply(dfSamples, 1, mean)

# dfSamples$xbarCumulative is appended as a column of cumulative means
dfSamples$xbarCumulative <- cumsum(dfSamples$xbar)/(1:numSimulations)

xbarTheoretical <- 1 / lambda  # the theoretical mean 

# Prepare plot of cumulative means, demonstrating convergence to the theoretical mean
meanPlot <- ggplot(data.frame(x = 1:numSimulations,
                              y = dfSamples$xbarCumulative),
            aes(x = x, y = y)) + geom_line() +
            geom_hline(aes(yintercept=xbarTheoretical,
                           color="red")) +
                geom_text(aes(750,xbarTheoretical,
                              label = "mean = 5",
                              vjust = 1,
                              hjust = 0)) +
            labs(title='Exponential Means: Empirical vs. Theoretical',
                 x='Simulations',y='Means')
```
```{r plotConvergence,echo=FALSE}
# display the plot
meanPlot
```


# Sample Variance versus Theoretical Variance

The theoretical variance of exponential samples is *(1/lambda) ^ 2*.  Computing the variance of the samples for the one thousand samples, I plot the rolling value of the variance as *n* increases.  

```{r compare_variance,}
# Compute variance for each sample of 40 exponentials, and append to  data.frame
dfSamples$varEmpirical <- apply(dfSamples[,paste("V",1:40,sep="")],1,var) 

# Compute a rolling estimate of variance as the number of samples increases.
dfSamples$varCumulative <- cumsum(dfSamples$varEmpirical)/(1:numSimulations)

# compute the theoretical variance
varTheoretical <- (1/lambda) ^ 2 

# Prepate plot 
varPlot <- ggplot(data.frame(x = 1:numSimulations,
                             y = dfSamples$varCumulative),
            aes(x = x, y = y)) +
            geom_line() +
            geom_hline(aes(yintercept=varTheoretical,
                           color="red")) +
                geom_text(aes(750,varTheoretical,
                              label = "var = 25",
                              vjust = 1,
                              hjust = 0)) +
            labs(title='Exponential Variance: Empirical vs. Theoretical',
                 x='Simulations',y='Variance')

# diplay plot
varPlot
```

# The Distribution of the Means
We expect that the distribution of the means will be normal, centered around the theoretical mean.  

# Appendix: Code Samples & Plots

## Libraries
```{r displayLibraries, ref.label='libraries',eval=FALSE }

```
## Generating Data
In the code below R generates a 1000 samples of 40 exponentials with lambda value of 0.2, converting it immediately to a data.frame named *dfSamples*.
```{r displayIntroduction,ref.label='introduction',eval=FALSE}

```
## Convergence of the Mean

```{r displayCompare_Mean,ref.label='compare_mean',eval=FALSE}

```

