---
title: "Statistical Inference Course Project"
author: "Chris van Hasselt"
date: "Monday, May 11, 2015"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r libraries, echo=FALSE }
# libraries loaded here
library(ggplot2)

```

# Overview:

The exponential distribution can be simulated in R with *rexp(n, lambda)* where *lambda* is the rate parameter. The mean of exponential distribution is *1/lambda* and the standard deviation is also also *1/lambda*. In this report, I will investigate the distribution of averages of 40 exponentials, when the rate parameter *lambda = 0.2*.

Specifically, I illustrate the properties of the distribution of the mean of 40 exponentials, showing:

#. The sample mean and compare it to the theoretical mean of the distribution;
#. variability (via variance) with respect to the theoretical variance of the distribution;
#. that the distribution is approximately normal.

Plots generated in this document are created using the *ggplot2* library.


# Simulations:

To analyze empirical versus theoretical statistics, I will create a data.frame of simulated random exponential values based on the given rate, using R.  In the code below R generates a matrix with 1000 samples of 40 exponentials with lambda value of 0.2, converting it immediately to a data.frame named *dfSamples*.  The intermediate *sampleMatrix* is kept

```{r intro}

# lambda is the exponential rate
lambda <- 0.2

# numSimulations is the number of simulations to perform
numSimulations <-1000

# sampleSize is the number of exponential samples per simulation
sampleSize <- 40

# allSamples is list of numSimulations (1000) times sampleSize (40) exponentials, with rate lambda
allSamples <- rexp(sampleSize*numSimulations, rate=lambda)

# sampleMatrix packages the exponentials as a matrix with numSimulations (1000) rows and sampleSize (40) columns.
sampleMatrix <- matrix(allSamples, numSimulations)

# dfSamples is a data.frame of all the samples, with sampleSize columns, numSimulations rows.
dfSamples   <- as.data.frame(sampleMatrix)

```

# Sample Mean versus Theoretical Mean:

To compare the empirical sample mean to the the theoretical mean, I append row means to the data.frame *dfSamples* as *dfSamples$xbar*, and cumulative row means as *dfSamples$xbarCumulative*.

```{r comparison}

# dfSamples$xbar is an appended column of row means
dfSamples$xbar <- apply(dfSamples, 1, mean)

# dfSamples$xbarCumulative is an appended column of cumulative means
dfSamples$xbarCumulative <- cumsum(dfSamples$xbar)/(1:numSimulations)

# xBarTheoretical is the theoretical mean for an exponential distribution, as stated in the problem
xBarTheoretical <- 1 / lambda


```

# Sample Variance versus Theoretical Variance



# Distribution
