---
title: "Inference Course Project"
author: "Chris van Hasselt"
date: "Monday, October 20, 2014"
output:
  pdf_document:
    latex_engine: xelatex
---

# Problem:

The exponential distribution can be simulated in R with *rexp(n, lambda)* where *lambda* is the rate parameter. The mean of exponential distribution is *1/lambda* and the standard deviation is also also *1/lambda*. Set *lambda = 0.2* for all of the simulations. In this simulation, you will investigate the distribution of averages of 40 exponential(0.2)s. Note that you will need to do a thousand or so simulated averages of 40 exponentials.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponential(0.2)s.  You should

#. Show where the distribution is centered at and compare it to the theoretical center of the distribution.
#. Show how variable it is and compare it to the theoretical variance of the distribution.
#. Show that the distribution is approximately normal.
#. Evaluate the coverage of the confidence interval for 1/lambda: $\bar{X} \pm 1.96\frac{\mathit{s}}{\sqrt[]{n}}$.


# Solution:

Prior to answering any questions about the distribution, I will use R to create a matrix of simulated random exponential values based on the given rate.

```{r}
# lambda is the exponential rate
lambda <- 0.2

# numSimulations is the number of simulations to perform
numSimulations <-1000

# sampleSize is the number of exponential samples per simulation
sampleSize     <- 40

# sampleMatrix is a matrix of all the samples, with sampleSize columns, numSimulations rows.
sampleMatrix   <- matrix(rexp(numSimulations * sampleSize, rate=lambda), numSimulations)

```
## Question 1

To show where the distribution is centered and compare to the theortical center of distribution, I use R to compute the mean of means, *xBarEmpirical*.  *xBarTheoretical* is the defined as *1/lambda*, as stated in the problem.  To compare the  

```{r}

# xBarEmprical is the mean of all sample means
xBarEmpirical  <- mean(apply(sampleMatrix, 1, mean))

# xBarTheoretical is 1/lambda for an exponential distribution, as stated in the problem
xBarTheoretical <- 1 / lambda

xBarDifference <- xBarEmpirical - xBarTheoretical
xBarDiffPct <- xBarDifference / xBarTheoretical

```

