---
title: "Statistical Inference Course Project"
author: "Chris van Hasselt"
date: "Monday, May 11, 2015"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r libraries, echo=FALSE }
# libraries loaded here

```

# Overview:

The exponential distribution can be simulated in R with *rexp(n, lambda)* where *lambda* is the rate parameter. The mean of exponential distribution is *1/lambda* and the standard deviation is also also *1/lambda*. In this report, I will investigate the distribution of averages of 40 exponentials, when the rate parameter *lambda = 0.2*.

Specifically, I illustrate the properties of the distribution of the mean of 40 exponentials, showing:

#. The sample mean and compare it to the theoretical mean of the distribution;
#. variability (via variance) with respect to the theoretical variance of the distribution;
#. that the distribution is approximately normal.


# Simulations:

Prior to addressing questions about the distribution, I will create a matrix of simulated random exponential values based on the given rate, using R.  In the code sample below, R generates the matrix *sampleMatrix* with 1000 samples of size 40.

```{r intro}
# lambda is the exponential rate
lambda <- 0.2

# numSimulations is the number of simulations to perform
numSimulations <-1000

# sampleSize is the number of exponential samples per simulation
sampleSize     <- 40

# sampleMatrix is a matrix of all the samples, with sampleSize columns, numSimulations rows.
sampleMatrix   <- matrix(rexp(numSimulations * sampleSize, rate=lambda), numSimulations)


```


# Sample Mean versus Theoretical Mean:

To show where the distribution is centered and compare to the theortical center of distribution, I use R to compute the mean of means, *xBarEmpirical*.  *xBarTheoretical* is the defined as *1/lambda*, as stated in the problem.  To compare the  

```{r Comparison}

# xBarEmprical is the mean of all sample means
xBarEmpirical  <- mean(apply(sampleMatrix, 1, mean))

# xBarTheoretical is 1/lambda for an exponential distribution, as stated in the problem
xBarTheoretical <- 1 / lambda

xBarDifference <- xBarEmpirical - xBarTheoretical
xBarDiffPct <- xBarDifference / xBarTheoretical

```

