---
title: "Statistical Inference Course Project"
author: "Chris van Hasselt"
date: "Sunday, May 24, 2015"
output:
  pdf_document:
    fig_height: 3.75
    fig_width: 5
    latex_engine: xelatex
---
```{r libraries, echo=FALSE }
# libraries loaded here
library(ggplot2)
```
# Overview:
The exponential distribution is simulated in R with *rexp(n, lambda)* where *lambda* is the rate parameter. The mean of the exponential distribution is *1/lambda* and the standard deviation is *1/lambda*.  This report examines the distribution of averages of 40 exponentials with rate *lambda = 0.2*, using plots generated with the *ggplot2* library, illustrating important properties of the distribution:

#. The sample mean converges to the theoretical mean of the distribution;
#. The sample variance converges to the theoretical variance of the distribution;
#. The distribution is approximately normal.

# Simulations:
To analyze empirical versus theoretical statistics, I create a data.frame named *dfSamples* of random exponentials with rate *lambda = 0.2*.  Code available in appendices.

```{r introduction,echo=FALSE, warning=FALSE}

set.seed(2132)          # set random seed, for repeatability

lambda <- 0.2           # define lambda, the exponential rate

numSimulations <- 1000  # numSimulations = number of simulations (1000)

sampleSize <- 40        # sampleSize = number of exponentials per simulation (40)

# allSamples is the numSimulations (1000) times sampleSize (40) exponentials
allSamples <- rexp(sampleSize*numSimulations, rate=lambda)

# sampleMatrix packages the exponentials as a matrix 
sampleMatrix <- matrix(allSamples, numSimulations)

# dfSamples is a data.frame of all the samples
dfSamples   <- as.data.frame(sampleMatrix)

```

# Sample Mean versus Theoretical Mean:
Comparing the empirical mean to the theoretical mean, I calculate row means and cumulative row means for data.frame *dfSamples*.  The plot below shows the sample means of 40 exponentials converging to the theoretical mean of *1/lambda = 5*, demonstrating asymptotic convergence.
```{r compare_mean, echo=FALSE, warning=FALSE}
# dfSamples$xbar is a column appended of row means
dfSamples$xbar <- apply(dfSamples, 1, mean)

# dfSamples$xbarCumulative is appended as a column of cumulative means
dfSamples$xbarCumulative <- cumsum(dfSamples$xbar)/(1:numSimulations)

xbarTheoretical <- 1 / lambda  # the theoretical mean 

# Prepare plot of cumulative means, demonstrating convergence to the theoretical mean
meanPlot <- ggplot(data.frame(x = 1:numSimulations,
                              y = dfSamples$xbarCumulative),
            aes(x = x, y = y)) + geom_line() +
            geom_hline(aes(yintercept=xbarTheoretical,
                           color="red")) +
                geom_text(aes(750,xbarTheoretical,
                              label = "mean = 5",
                              vjust = 1,
                              hjust = 0)) +
            labs(title='Exponential Means: Empirical vs. Theoretical',
                 x='Simulations',y='Means')

# display the plot
meanPlot
```

# Sample Variance versus Theoretical Variance
Plotting the rolling value of the variance as *n* increases, we see it converging to the expected theoretical variance *(1/lambda) ^ 2 = 25* (red line), demonstrating asymptotic convergence of the variance.

```{r compare_variance,echo=FALSE, warning=FALSE}
# Compute variance for each sample of 40 exponentials, and append to  data.frame
dfSamples$varEmpirical <- apply(dfSamples[,paste("V",1:40,sep="")],1,var) 

# Compute a rolling estimate of variance as the number of samples increases.
dfSamples$varCumulative <- cumsum(dfSamples$varEmpirical)/(1:numSimulations)

# compute the theoretical variance
varTheoretical <- (1/lambda) ^ 2 

# Prepate plot 
varPlot <- ggplot(data.frame(x = 1:numSimulations,
                             y = dfSamples$varCumulative),
            aes(x = x, y = y)) +
            geom_line() +
            geom_hline(aes(yintercept=varTheoretical,
                           color="red")) +
                geom_text(aes(750,varTheoretical,
                              label = "var = 25",
                              vjust = 1,
                              hjust = 0)) +
            labs(title='Exponential Variance: Empirical vs. Theoretical',
                 x='Simulations',y='Variance')

# diplay plot
varPlot
```
\pagebreak

# The Distribution of the Means
Plotting the histogram with its empirical density overlaying (pink) shows the distribution is approximately normal, centered around the theoretical mean.  The normal distribution (orange) is also shown.
```{r mean_distribution,echo=FALSE}
# display a plot of the histogram;
xbarDist <- ggplot(data=dfSamples, aes(dfSamples$xbar)) + 
                geom_histogram(aes(y=..density..),
                               col="darkblue",fill="grey",
                               binwidth=diff(range(dfSamples$xbar))/30) +
                geom_density(alpha=.2, fill="#FF6666") + 
                geom_vline(aes(xintercept=xbarTheoretical,
                           color="red")) +
                geom_text(aes(xbarTheoretical,.5,
                              label = "mean = 5",
                              vjust = 1,
                              hjust = 1.2)) +
                stat_function(fun = dnorm, arg=list(mean=xbarTheoretical), 
                              colour = "#FFCC33",
                              size=1.2) +
                labs(title='Distribution of Means with Expected Normal Curve',
                     x='Computed Means',
                     y='Density')
# plot the distribution
xbarDist
```
\pagebreak

# Appendices: Code Samples & Plots

## Libraries
```{r display_libraries, ref.label='libraries',eval=FALSE }

```
## Generating Data
In the code below R generates a 1000 samples of 40 exponentials with lambda value of 0.2, converting it immediately to a data.frame named *dfSamples*.
```{r display_introduction,ref.label='introduction',eval=FALSE}

```
## Convergence of the Mean
The code below computes the means for each row of 40 samples, and prepares the plot demonstrating asymptotic convergence of the mean.
```{r display_compare_mean,ref.label='compare_mean',eval=FALSE}

```
## Convergence of the Variance
The code below is computes the variance for each row of 40 samples, and prepares the plot demonstrating asymptotic convergence of the mean.
```{r display_compare_variance,ref.label='compare_variance',eval=FALSE}

```

## The Distribution of the Means
The code below creates the plot of the empirical mean as a histogram, with a density function overlaying it, and the true normal distribution displayed as well. The plot shows that the empirical distribution is approximately normal.
```{r display_mean_distribution,ref.label='mean_distribution',eval=FALSE}

```